
```{r setup, cache = FALSE, echo = FALSE}
# load required libraries
library(fda)
library(lattice)

dir <- "B:/LochArd_Fellng/"

opts_chunk $ set(fig.width = 10, fig.height = 10, 
                 fig.path   = paste0(dir, "Rmd/simulation/fig/"),
                 cache = FALSE,
                 cache.path = paste0(dir, "Rmd/simulation/cache/"),
                 echo = FALSE)
options(digits = 5)
```

# Simulating river temperature

We are going to simulate some (probably wierd looking) temperature time series.  The purpose is to investigate what the model can do.  Lets keep things simple first off and use 4 degrees of freedom for the hourly smoothers and 10 degrees of freedom for the yearly smoothers.

```{r outline, echo = TRUE}
# number of parameters in hourly model
ph <- 6
# number of parameters in daily model
pd <- 10

# sample points
hours <- seq(0, 24, by = 0.25)
days <- 1:365

# lets use a bspline basis for both
hourbasis <- create.bspline.basis(c(0, 24), ph, 6)
hourX <- Matrix(getbasismatrix(hours, hourbasis), sparse = TRUE)

daybasis <- create.bspline.basis(c(1, 365), pd, 6)
dayX <- Matrix(getbasismatrix(days, daybasis), sparse = TRUE)

# the full X matrix is
X <- Diagonal(365) %x% hourX

# the full Z matrix is
Z <- Diagonal(ph) %x% dayX

# The permutation matrix is
P <- Matrix(0, 365*ph, 365*ph)
P[1:365,] <- Diagonal(365) %x% Matrix(c(1, rep(0, ph-1)), 1)
for (i in 1:(ph-1)) P[i * 365 + 1:365,] <- Diagonal(365) %x% Matrix(c(rep(0, i), 1, rep(0, ph-1-i)), 1)

# so the full model (without AR1 noise on the coeffients is)
A <- X %*% P %*% Z

# lets choose some random coefficients and plot what we get
alpha <- rnorm(ph*pd, 0, 1)

df <- expand.grid(hour = hours, day = days)
df $ month <- factor(month.abb[floor(df $ day / 30.5) + 1], month.abb)
df $ model <- A %*% alpha

xyplot(model ~ hour | month, group = day, data = df, type = "l")
```

okay so it looks daft, so lets parameterise it a bit better after conducting a PCA on daily temperatures

```{r pca, cache = TRUE}
# get full data set
load("B:/Loch_Ard_Felling/Package/rivertemp/data/fullwater.rda")
# a quick fix
stream $ year[stream $ Sensor.Model == "Squirrel"] <- stream $ year[stream $ Sensor.Model == "Squirrel"] + 1900 
# set up a bspline basis of dimension 48 and order 6 
hourbasis <- create.bspline.basis(c(0, 24), 48, 6)
# and a harmonic accelerator penalty... penalises deviations from sinusiodal curves
penalty <- vec2Lfd(c(0, (2*pi/24)^2, 0), c(0, 24))

# select a few days data and add in decimal hour
dat <- with(subset(stream, site == 2 & year %in% 2003:2003), 
    data.frame(temp = Original.Value, 
               dhour = hour + min/60, 
               month = factor(month.abb[mon + 1], levels = month.abb), 
               yday = yday, 
               year = year, 
               site = site))
dat <- subset(dat, yday < 365) # remove last day of leap year

getcoef <- function(lambda, .yday, .year = 2009, .site = 2) {
  sdat <- subset(dat, yday == .yday & year == .year & site == .site)
  #if (nrow(sdat) < 23) return (NULL)
  if (sum(unique(round(sdat $ dhour)) %in% 0:24) < 24) return(NULL)  
  c(coef(smooth.basis(sdat $ dhour, sdat $ temp, fdPar(hourbasis, penalty, lambda))))
}

yday <- rep(0:365, length(2003:2009))
year <- rep(2003:2009, each = 366)

ck <- mapply(getcoef, .yday = yday, .year = year, MoreArgs = list(lambda = exp(-8)))
whichNULL <- sapply(ck, is.null)
ck <- simplify2array(ck[!whichNULL])

tempfd2 <- fd(ck, hourbasis)
pc2 <- pca.fd(tempfd2, nharm = 4) # is this sensitive to missing values?...
par(mfrow = c(2,2))
plot(pc2)
```

```{r newbasis_fd}
newbasis <- pc2 $ harmonics
plot(newbasis)
```

try and simulate some days based on these new bases

```{r outline2, echo = TRUE}
# number of parameters in hourly model
ph <- 4 # by design
# number of parameters in daily model
pd <- 10

# sample points
hours <- seq(0, 24, by = 0.25)
days <- 1:365

# lets use a bspline basis for both
hourX <- Matrix(eval.fd(hours, newbasis), sparse = TRUE)

daybasis <- create.bspline.basis(c(1, 365), pd, 6)
dayX <- Matrix(getbasismatrix(days, daybasis), sparse = TRUE)

# the full X matrix is
X <- Diagonal(365) %x% hourX

# the full Z matrix is
Z <- Diagonal(ph) %x% dayX

# The permutation matrix is
P <- Matrix(0, 365*ph, 365*ph)
P[1:365,] <- Diagonal(365) %x% Matrix(c(1, rep(0, ph-1)), 1)
for (i in 1:(ph-1)) P[i * 365 + 1:365,] <- Diagonal(365) %x% Matrix(c(rep(0, i), 1, rep(0, ph-1-i)), 1)

# so the full model (without AR1 noise on the coeffients is)
A <- X %*% P %*% Z

# lets choose some random coefficients and plot what we get
alpha <- rnorm(ph*pd, 0, 1)

df <- expand.grid(hour = hours, day = days)
df $ month <- factor(month.abb[floor(df $ day / 30.5) + 1], month.abb)
df $ model <- A %*% alpha

xyplot(model ~ hour | month, group = day, data = df, type = "l")
``

now down to 40 parameters though!  But we could do better by doing a PCA on each basis coefficients time series.  Lets try and improve the 1st PC.

```{r }

```

